---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      collapse = TRUE,
                      comment = "#>",fig.width = 10)
```

## Purpose: 
Showing correlations of samples within and between methods as well as comparability of dat sets.

### 1. Load the following packages:
```{r}
library(ggplot2)
library(cowplot)
library(tidyverse)
library(corrplot)
library(RColorBrewer)
library(ggbeeswarm)
library(UpSetR)
library(ggplotify)
```


```{r}
### all necessary custom functions are in the following scripts
source(paste0(here::here(),"/0_Scripts/custom_functions.R"))

theme_pub <- theme_bw() + theme(plot.title = element_text(hjust = 0.5, 
                                                          size=18, face="bold"),
                                     axis.text = element_text(colour="black", size=14), 
                                     axis.title=element_text(size=16,face="bold"), 
                                     legend.text=element_text(size=14),
                                     legend.position="right",
                                     axis.line.x = element_line(colour = "black"), 
                                     axis.line.y = element_line(colour = "black"),
                                     strip.background=element_blank(), 
                                     strip.text=element_text(size=16))  
theme_set(theme_pub)


fig_path <- paste0(here::here(),"/2_power_simulation/")


## colours

method_cols<-c("#008080","gray70","#457b9d")
names(method_cols)<-c("prime-seq","TruSeq","between")

endo_cols<-c("#19294D","#8F8F8F","#3E4451")
names(endo_cols)<-c("Spike-Ins","Endogenous Genes","Endogenous Genes2")

```

### 2. Load the data
```{r}

## tru_seq

tru_inex<-readRDS(paste0(fig_path,"/SEQC_PE.dgecounts.rds"))$readcount$inex$downsampling$downsampled_20000000 %>% as.matrix()

tru_ercc<-tru_inex[grep(rownames(tru_inex),pattern="ERCC*"),]
tru_inex<-tru_inex[grep(rownames(tru_inex),pattern="ERCC*",invert = T),]
tru_inex<-remove_Geneversion(as.matrix(tru_inex))
tru_inex<-tru_inex[whichgenes_reproducible(tru_inex,exprcutoff = 1,reproducecutoff = 0.25),]
tru_inex<-rbind(tru_inex,tru_ercc)

## prime_seq

# Reads
prime_inex_reads<-readRDS(paste0(fig_path,"/prime-seq.dgecounts.rds"))$readcount$inex$downsampling$downsampled_20000000 %>% as.matrix()


prime_ercc_reads<-prime_inex_reads[grep(rownames(prime_inex_reads),pattern="ERCC*"),]
prime_inex_reads<-prime_inex_reads[grep(rownames(prime_inex_reads),pattern="ERCC*",invert = T),]
prime_inex_reads<-remove_Geneversion(as.matrix(prime_inex_reads))
prime_inex_reads<-prime_inex_reads[whichgenes_reproducible(prime_inex_reads,exprcutoff = 1,reproducecutoff = 0.25),]
prime_inex_reads<-rbind(prime_inex_reads,prime_ercc_reads)


# UMIs
prime_inex<-readRDS(paste0(fig_path,"/prime-seq.dgecounts.rds"))$umicount$inex$downsampling$downsampled_20000000 %>% as.matrix()

prime_ercc<-prime_inex[grep(rownames(prime_inex),pattern="ERCC*"),]
prime_inex<-prime_inex[grep(rownames(prime_inex),pattern="ERCC*",invert = T),]
prime_inex<-remove_Geneversion(as.matrix(prime_inex))
prime_inex<-prime_inex[whichgenes_reproducible(prime_inex,exprcutoff = 1,reproducecutoff = 0.25),]
prime_inex<-rbind(prime_inex,prime_ercc)




### Combining dataframes
comb_inex<-merge(tru_inex,prime_inex,by="row.names",all=F)

comb_inex_reads<-merge(tru_inex,prime_inex_reads,by="row.names",all=F)


comb_inex<-comb_inex %>% 
  column_to_rownames(var = "Row.names") %>% 
  as.matrix()


comb_inex_reads<-comb_inex_reads %>% 
  column_to_rownames(var = "Row.names") %>% 
  as.matrix()


### make info
info<-data.frame(BC=c(colnames(prime_inex),colnames(tru_inex)),
                 method=rep(c("prime-seq","TruSeq"),c(8,5)))

```
### 3. Normalize
```{r}
library(DESeq2)
des_reads <- DESeqDataSetFromMatrix( countData = comb_inex_reads,
                               colData = info,
                               design = ~ 0 + method)

des_reads<-DESeq2::varianceStabilizingTransformation(des_reads)

norm_reads<-assay(des_reads)

library(DESeq2)
des_umi <- DESeqDataSetFromMatrix( countData = comb_inex,
                               colData = info,
                               design = ~ 0 + method)

des_umi<-DESeq2::varianceStabilizingTransformation(des_umi)

norm_umi<-assay(des_umi)

```

### 4. Pairwise correlations within and between methods
```{r}
log_mat<-norm_umi

log_mat_reads<-norm_reads

lt <- cor(log_mat,method = "pearson")

lt <-lt^2
  
lt[lower.tri(lt)]<-""
diag(lt)<-""

cor_df<-lt %>% 
  reshape2::melt(value.name = "cor",varnames=c("BC", "BC2")) %>%
  filter(cor!="") %>% 
  left_join(info) %>% 
  select(-BC) %>% 
  left_join(info,by=c("BC2"="BC")) %>% 
  select(-BC2) %>% 
  mutate(type=if_else(method.x==method.y,"within","between"),
         method_col=if_else(type=="between","between",method.x)) %>% 
  mutate(cor=as.numeric(cor),
         rr=as.numeric(cor))

mean_df<-cor_df %>% 
  group_by(method_col) %>% 
  summarise(cor=round(median(cor),digits = 2))

sample_cor_umi<-ggplot(cor_df,aes(x=method_col,y=cor,colour=method_col))+
  geom_beeswarm()+
  geom_label(data=mean_df,aes(label=cor,y=0.55,fill=method_col),colour="white")+
  scale_colour_manual(values = method_cols)+
  scale_fill_manual(values = method_cols)+
  labs(x="",y=bquote("pairwise Pearson's R"^2))+
  theme(legend.position = "none",
        axis.text.x=element_text(size=16,face="bold"))+
  scale_y_continuous(limits = c(0.5,1))

```

### 5. Correlate mean expression

```{r}


mean_exp_umi<-log_mat%>% 
  as.data.frame() %>% 
  rownames_to_column(var="Gene") %>% 
  pivot_longer(names_to = "BC",values_to="count",cols=-Gene) %>% 
  left_join(info) %>% 
  group_by(method,Gene) %>% 
  dplyr::summarize(mean_exp=mean(count)) %>% 
  mutate(spike_in=if_else(Gene %in% grep(Gene,pattern="ERCC*",
                                         value = T),
                          "Spike-Ins","Endogenous Genes")) %>% 
  pivot_wider(names_from = "method",values_from="mean_exp") %>% 
  as.data.frame()

mean_exp_umi<-mean_exp_umi %>% 
  mutate(spike_label=if_else(spike_in=="Endogenous Genes","Endogenous Genes2",spike_in))

endo<-subset(mean_exp_umi,spike_in=="Endogenous Genes")
spike<-subset(mean_exp_umi,spike_in!="Endogenous Genes")

gene_cor_umi<-ggplot()+
  geom_point(data=endo,
             aes(y=`prime-seq`,
                 x=`TruSeq`,
                 colour=spike_in),
             alpha=0.2,
             size=0.2)+
  geom_point(data=spike,
             aes(y=`prime-seq`,
                 x=`TruSeq`,
                 colour=spike_in),
             alpha=1,
             size=1)+
  geom_smooth(data=mean_exp_umi,
              aes(y=`prime-seq`,
                  x=`TruSeq`,
                  colour=spike_label),
              method = "lm",
              show.legend = F)+
  ggpubr::stat_cor(data=mean_exp_umi,
                   aes(y=`prime-seq`,
                       x=`TruSeq`,
                       fill=spike_in,
                       label=paste(..rr.label..)),
                   method = "pearson",
                   geom = "label",
                   colour="white",
                   label.x = c(2),
                   label.y=c(12,14.5),
              show.legend = F)+
  theme(axis.line.y = element_line(),
        legend.position="right")+
  scale_colour_manual(values=endo_cols,limits= force )+
  scale_fill_manual(values=endo_cols)+
  xlab("prime-seq \n normalized expression ")+
  ylab("TruSeq \n normalized expression")+
  labs(colour="")+
  ylim(0,15)+
  xlim(0,15)


gene_cor_umi

plot_grid(sample_cor_umi,gene_cor_umi,nrow=1)

```



### 6. upsetR plot of genes expressen in TruSeq and prime-seq
```{r}

comb_inex_full<-merge(tru_inex,prime_inex,by="row.names",all=T)

upset_df<-data.frame(prime=rowSums(comb_inex_full[,colnames(comb_inex_full)%in%subset(info,method=="prime-seq")$BC]),
                        TruSeq=rowSums(comb_inex_full[,colnames(comb_inex_full)%in%subset(info,method=="TruSeq")$BC]),
                        Gene=rownames(comb_inex_full))

upset_df <- upset_df %>%
  mutate(prime = ifelse(prime > 0, paste(Gene), NA),
         TruSeq = ifelse(TruSeq > 0, paste(Gene), NA)) %>%
  dplyr::select(-Gene) %>%
  filter_all(any_vars(!is.na(.)))

upset_comp <- upset(fromList(upset_df), order.by = "freq",
                 mainbar.y.label = "Gene Intersections", 
                 sets.x.label="Genes per Condition",
                 main.bar.color = "#457b9d", 
                 sets.bar.color = "#14213d", 
                 text.scale = c(2, 2, 2, 2, 2, 2),
                 point.size = 3.5,show.numbers = F)

upset_comp

table(fromList(upset_df))

```


### 7. Plot mean dispersion plot for both methods
```{r}
load(paste0(fig_path,"/estparam_all.Rdata"))

mv_prime<-Get_Mean_Disp(estparam_gene_prime_inex)

mv_tru<-Get_Mean_Disp(estparam_gene_tru_inex)

meanvsdisp.dat<-rbind(mv_prime$meanvsdisp.dat,mv_tru$meanvsdisp.dat)
meanvsdisp.dat$method<-rep(c("prime-seq","TruSeq"),times=c(nrow(mv_prime$meanvsdisp.dat),nrow(mv_tru$meanvsdisp.dat)))

meanvsdisp.fdat<-rbind(mv_prime$meanvsdisp.fdat,mv_tru$meanvsdisp.fdat)
meanvsdisp.fdat$method<-rep(c("prime-seq","TruSeq"),times=c(nrow(mv_prime$meanvsdisp.fdat),nrow(mv_tru$meanvsdisp.fdat)))

cdisp<-c(mv_prime$cdisp,mv_tru$cdisp)

Fig_mean_disp <- ggplot(data=meanvsdisp.dat,
                                     aes(x=Mean,
                                         y=Dispersion,
                                         col=method)) +
  geom_point(size=0.5,alpha=0.1) +
  geom_line(data = meanvsdisp.fdat,
                      aes(x = Mean,
                           y = Dispersion,
                           col=method),
                      linetype=1,
                      size=1.5) +
  geom_line(data = meanvsdisp.fdat,
                       aes(x = Mean,
                           col=method),
              linetype=2,
              size=1) +
  geom_line(data = meanvsdisp.fdat,
                       aes(x = Mean,
                            y = Upper,
                           col=method),
                       linetype=2,
                       size=1) +
   geom_line(data = meanvsdisp.fdat,
                       aes(x = Mean,
                            y = Lower,
                           col=method),
                       linetype=2,
                       size=1) +
  geom_abline(intercept = 0,
                        slope = -1,
                        linetype = 2,
                        colour="grey40",
                        size=1.5) +
  scale_colour_manual(values=method_cols[1:2])+
  labs(x="Log2 Mean Expression",
        y="Log2 Dispersion",colour="")+
  theme(legend.position="bottom")+
  theme(legend.position = "right")
 
Fig_mean_disp_mar<-Fig_mean_disp+scale_x_continuous(limits = c(0,12))
Fig_mean_disp_mar<-ggExtra::ggMarginal(Fig_mean_disp_mar, type="histogram", 
                                       size=2,margins = "x",groupFill = T)

Fig_mean_disp_rev<-ggExtra::ggMarginal(Fig_mean_disp,
                                       type="histogram",
                                       size=2,
                                       margins = "x",
                                       groupFill = T)


Fig_mean_count<-ggplot(data=meanvsdisp.dat,aes(x=Mean))+
  geom_histogram(aes(fill=method),bins = 100)+
  scale_fill_manual(values=method_cols[1:2])+
  labs(x="Log2 Mean Expression",y="Number of Genes")+
  theme(legend.position = "none")+
  facet_grid(method~.)
  
Fig_3B<-plot_grid(Fig_mean_count,Fig_mean_disp,ncol=2)

Fig_3B

Fig_mean_disp_rev
```

### 8. Save plots

```{r}

# Exporting Individual Plots

ggsave(Fig_mean_disp_mar,
       device = "pdf",
       path = fig_path,
       width = 200,
       height=95,
       units = "mm",
       filename = "Fig3_Mean_disp.pdf"
       )

ggsave(Fig_mean_disp_mar,
       device = "png",
       path = fig_path,
       width = 200,
       height=95,
       units = "mm",
       filename = "Fig3_Mean_disp.png"
       )

ggsave(Fig_mean_disp_rev,
       device = "png",
       path = fig_path,
       width = 200,
       height=95,
       units = "mm",
       filename = "Fig3_Mean_disp_rev.png"
       )

ggsave(as.grob(upset_comp),
       device = "pdf",
       path = fig_path,
       width = 170,
       height= 130,
       units = "mm",
       filename = "Tru_prime_upset.pdf"
       )


ggsave(sample_cor_umi,
       device = "pdf",
       path = fig_path,
       width = 105,
       height= 90,
       units = "mm",
       filename = "Tru_prime_paircor.pdf"
       )


ggsave(gene_cor_umi,
       device = "png",
       path = fig_path,
       width = 210,
       height= 95,
       units = "mm",
       filename = "Tru_prime_cor.png"
       )


```

## `R` Session Info

```{r}
sessionInfo()
```